---
# IMPORTANT: Change settings here, but DO NOT change the spacing.
# Remove comments and add values where applicable.
# The descriptions below should be self-explanatory

title: "Data Science Practical"
#subtitle: "Question 1-5"

documentclass: "elsarticle"
"
# Entry5: "Stellenbosch University"
# Entry6: June 2022

# --------- Front Page


# --------- page margins:
margin: 2.3 # Sides
bottom: 2 # bottom
top: 2.5 # Top
HardSet_layout: TRUE # Hard-set the spacing of words in your document. This will stop LaTeX squashing text to fit on pages, e.g.
# This is done by hard-setting the spacing dimensions. Set to FALSE if you want LaTeX to optimize this for your paper.


### Adding additional latex packages:
# header-includes:
#    - \usepackage{colortbl} # Add additional packages here.

output:
  pdf_document:
    keep_tex: TRUE
    template: Tex/TexDefault.txt
    fig_width: 3.5 # Adjust default figure sizes. This can also be done in the chunks of the text.
    fig_height: 3.5
abstract: |

---

<!-- First: Set your default preferences for chunk options: -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.pos="H", fig.pos = 'H')

```


```{r}
#Folder and follow up question folder created using the following code

Texevier::create_template(directory = "C:/Users/ngane/OneDrive/Documents/directory/26682729",
                          template_name = "26682729",
                          build_project = T, open_project = T)
#C:\Users\ngane\OneDrive\Documents\directory\26682729

Texevier::create_template(directory = glue::glue("directory/26682729/"), template_name = "Question1")
Texevier::create_template(directory = glue::glue("directory/26682729/"), template_name = "Question2") 
Texevier::create_template(directory = glue::glue("directory/26682729/"), template_name = "Question3") 
Texevier::create_template(directory = glue::glue("directory/26682729/"), template_name = "Question4") 
Texevier::create_template(directory = glue::glue("directory/26682729/"), template_name = "Question5") 

```


# Introduction \label{Introduction}

Question 1: Covid 
Using variables like total cases, total deaths and total number of people fully vaccinated 
```{r}
#wrangling data 
plotcovid<-
    select(owid.covid.data,location,continent,date,people_fully_vaccinated,total_cases,total_deaths,population) %>% 
    group_by(continent) %>% 
    mutate(casepop= total_cases/population) %>% 
    mutate(deathpop=total_deaths/population) %>%
    mutate(vacpop=people_fully_vaccinated/population) %>% ungroup()
print(plotcovid)
#plots for aforementioned variables
g <- plotcovid %>% 
ggplot() + 
geom_line(aes(x = date, y = casepop, color = continent), alpha = 0.1, 
    size = 1)
print(g)
g <- g + facet_wrap(~continent,scales = "free_y") + ggthemes::scale_color_wsj()

g
#total deaths by continent
h<- plotcovid %>% 
ggplot() + 
geom_line(aes(x = date, y = deathpop, color = continent), alpha = 0.1, 
    size = 1)
print(g)
h<- h+ facet_wrap(~continent,scales = "free_y") + ggthemes::scale_color_wsj()

h
#vaccination
d <- plotcovid %>% 
ggplot() + 
geom_line(aes(x = date, y = vacpop, color = continent), alpha = 0.1, 
    size = 1)
print(g)
d <- d + facet_wrap(~continent,scales = "free_y") + ggthemes::scale_color_wsj()

d

```
Second part of question 1
```{r}
library(dplyr)

Covidlex <- owid.covid.data %>%
  select(life_expectancy, continent, total_deaths) %>%
  group_by(continent,total_deaths) %>% 
  summarise(mean_life = mean(life_expectancy),
            LifeDiff = life_expectancy- mean_life,
            criteria = ifelse(LifeDiff > 15, "high", ifelse(LifeDiff < 15, "regular", "low"))) %>%
   ungroup()
# Print the Covidlex tibble
print(Covidlex)

# Filter the high values and select location and total_deaths
high_values <- Covidlex %>%
  filter(criteria == "high") %>%
  select(continent,total_deaths)

# Print the high_values tibble
print(high_values)

```
The findings are quite strange as the 'continent' variable as it is is only picking up Africa. We would expect to find more variation in the continents 
that have life expectancy greater than mean life expectancy.
```{r}
Covidlex%>% pull(continent ) %>% unique()
```
Confirming that all the variables of continent are in our smaller data frame
and they are.
Based on the findings, we can conclude that countries that have higher life
expectancy in Africa did not experience higher total deaths as their maximum value
was approximately 3000 even by the end of the pandemic and when compared to the
numbers reached by other countries even in the beginning it means they experienced
less loss in general.
```{r}
#increasing hospitalization facilities I used hospital_beds_per thousand and for
#checking lagged ICU i used ICU patients
library(dplyr)

covid_data <- owid.covid.data %>%
  select(continent, date, hospital_beds_per_thousand, icu_patients) 

library(dplyr)
library(ggplot2)
covid_data <- covid_data %>%
  mutate(icu_lag = icu_patients - lag(icu_patients))


covid_data <- covid_data %>%
  mutate(date = as.Date(date),  # Convert date column to proper date format
         hospital_beds_increase = hospital_beds_per_thousand - lag(hospital_beds_per_thousand)) %>% 

        mutate( hospital_beds_increase_rate =(hospital_beds_increase / hospital_beds_per_thousand)*100)

#plotting the two together

ggplot(covid_data, aes(x = date)) +
  geom_line(aes(y = hospital_beds_increase_rate, color = "Hospital Beds Increase"), size = 0.5) +
  geom_line(aes(y = icu_lag * 10, color = "ICU Admissions"), size = 0.5) +
  scale_color_manual(values = c("Hospital Beds Increase" = "green", "ICU Admissions" = "pink")) +
  labs(x = "Date", y = "Increase", color = "Variable") +
  theme_minimal() + scale_y_continuous()+
  theme(legend.position = "top") +
  facet_wrap(~ continent, ncol = 2,scales = "free_y")
```
From what we can see, Asia and Africa experienced increased rates of in their 
hospitalisation facilities while for the others it was constant.We can observe 
that Europe and South America went through tumultuous ICU lags even though we
do not observe increase in facilities. 

Question 2 
```{r}
#focus on 3 metrics here, sunny, snow, low temperatures as x in plots and y as dates (over time)
#DSND= number of days with snow abover 25mm (not in dataset)
#DX70= number of days with temperature below 21.1 degrees
#TSUN= Monthly total sunshine in minutes
#can't find most of these in dataset

UKtemp<- UKMonthly_Detailed %>% 
     mutate(date = as.Date(paste0(DATE, "-01"), format = "%Y-%m-%d")) %>%
    select(DX70,date)

# Plotting number of days with temperature below 21.1 degrees Celsius over time
ggplot(UKtemp, aes(x = date, y = DX70)) +
  geom_line(color = "lightblue") +
  labs(x = "Date", y = "Number of Days",
       title = "Number of Days with Temperature Below 21.1°C in Central England")
```
 
```{r}
#plotting how sunny and rainy its been in London over time to note any changes
#wrangling first
londonsun_rain<- london_weather %>% 
    select(date,precipitation,sunshine) %>% 
    mutate(date= as.Date(date, format="%Y-%m-%d", origin=("1970-01-01"))) %>% 
      filter(!is.na(date)) 
#issues with getting date from london_weather into correct format, histogram instead

ggplot(londonsun_rain, aes(x = sunshine)) +
  geom_histogram(color = "black", fill = "orange", bins = 30) +
  labs(x = "Sunny Hours", y = "Frequency",
       title = "Distribution of Sunny Hours in London") +
  theme_minimal()

ggplot(londonsun_rain, aes(x = precipitation)) +
  geom_histogram(color = "black", fill = "lightgreen", bins = 30) +
  labs(x = "Precipitation", y = "Frequency",
       title = "Distribution of Precipitation in London") +
  theme_minimal()
```
 So far, it is evident that the number of days with temperatures below 21.1 degrees celsius has increases somewhat in recent years. For the histograms of London weather, it is evident London does not get a lot of sunshine, with the highest probability being that of little to no sunshine.There's also not much rain that falls but it is for sure cold and dreary up there more often than not.
 
 Question 3
```{r}
#comparing coldplay and metallica, we begin with coldplay
coldplay_albums <- 
    Coldplay %>% select(album_name, release_date, name, popularity) %>%       group_by(album_name) %>% arrange(release_date) %>% 
    ungroup()

coldplay_plotdf <- 
    coldplay_albums %>% 
    group_by(album_name) %>% mutate(mean_pop = round(mean(popularity))) %>% 
    arrange(release_date) %>% mutate(popularity = as.numeric(popularity)) #%>%          plot_orderset(., Column = "album_name", Order = order1)

#ordering: 
# order1 <- coldplay_plotdf %>% arrange(release_date) %>% pull(album_name)
# 
# coldplay_plotdf$album_name <- factor(coldplay_plotdf$album_name, levels = order1)


b <- 
    coldplay_plotdf %>% 
    #plot_orderset(., Column = "album_name", Order = order1) %>% 
    ggplot() + 
    geom_bar(aes(album_name, popularity, fill = mean_pop), stat = 'identity') +
    scale_y_continuous(limits = c(0,100), breaks = scales::breaks_pretty(n = 10)) +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(x = "Names of Albums", y = "Popularity", caption = "Sourced: From Spotify database", title = "Coldplay's Avg. Ratings per album")

b

rmsfuns::build_path(paste0(getwd(), "Figures"))

ggsave(filename = "Figures/Plot.png", plot = b, width = 12, height = 10, 
    device = "png")
```
 On to Metallica
```{r}
#metallica stats
metallica_albums <- 
    metallica %>% select(album, release_date, name, popularity) %>%       group_by(album) %>% arrange(release_date) %>% 
    ungroup() 

met_plotdf<- 
    metallica_albums %>% 
    group_by(album) %>% mutate(mean_pop = round(mean(popularity))) %>% 
    arrange(release_date) %>% mutate(popularity = as.numeric(popularity)) %>%  mutate (album = as.factor(album)) %>% 
    mutate(album = fct_reorder(name, release_date)) #%>%          plot_orderset(., Column = "album_name", Order = order1)

#ordering: 
# order1 <- album_plotdf %>% arrange(release_date) %>% pull(album_name)
# 
# met_plotdf$album_name <- factor(met_plotdf$album_name, levels = order1)


c <- 
    met_plotdf %>% 
    #plot_orderset(., Column = "album_name", Order = order1) %>% 
    ggplot() + 
    geom_bar(aes(album, popularity, fill = mean_pop), stat = 'identity') +
    scale_y_continuous(limits = c(0,100), breaks = scales::breaks_pretty(n = 10)) +
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +

labs(x = "Names of Albums", y = "Popularity", caption = "Sourced: From Spotify database", title = "Metallica's Avg. Ratings per album")

c

rmsfuns::build_path(paste0(getwd(), "Figures"))

ggsave(filename = "Figures/Plot.png", plot = b, width = 12, height = 10, 
    device = "png")

```
 Question 4

```{r}
Movies  %>% 
mutate(Genre = forcats::fct_infreq(Genre)) %>% 
ggplot() + 
geom_bar(aes(x = Genre), fill = "lightgreen", alpha = 0.7) + 
    
coord_flip() + 
labs(x = "Count", y = "Genre", title = "Genre Spread of movies", 
    caption = "Data from imdmb")

```
    
```{r}
Movies %>% pull(Genre ) %>% unique()
titles %>% pull(genres ) %>% unique()
```
```{r}
```{r}
#plotting the genres against profits and then audience scores
#profits scaled in first figure due to outlier in drama throwing entire figure off
 
Movies <- Movies %>%
  select(Genre, Profitability, Audience..score..) %>%
  na.omit()

#filtering outlier in drama first (its a movie called Fireproof)
#filtered_movies <- Movies %>%
#filter(!(Genre == "Drama" & Profitability < 40))
#boxplot still does not look right, would need to keep pruning outliers. Will just adjust scale of y
  
ggplot(Movies, aes(x = Genre, y = Profitability, fill = Genre)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Genre", y = "Profitability", fill = "Genre")+  scale_y_continuous(trans = "log10")

```

Interpretations for profits no longer direct but can still discern that romance, drama and comedy are the most profitable. Action and fanstasy collapse due to the transformation


```{r}
ggplot(Movies, aes(x = Genre, y = Audience..score.., fill = Genre)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Genre", y = "AudienceScore", fill = "Genre")


```

```
 
